Speed and Precision in Range Analysis

### 2012-06-30 doc/Manuscript/SPE/paper.tex
Correção
Applied the reviews to the paper.

% The context: why range analysis is necessary.
The analysis of integer variables on the interval lattice has been the
canonical example of abstract interpretation since its introduction in
Cousot and Cousot's seminal paper~\cite{Cousot77}.
Compilers use range analysis to infer the possible values that discrete
variables may assume during program execution.
This analysis has many uses.
For instance, it allows the optimizing compiler to remove from the program text
redundant overflow tests~\cite{Sol11} and unnecessary array bound
checks~\cite{Bodik00,Gampe11}.
Furthermore, range analysis is essential not only to the bitwidth aware
register allocator~\cite{Barik06,Tallam03}, but also to more traditional
allocators that handle registers of different
sizes~\cite{Kong98,Pereira08,Scholz02}.
Additionally, range analysis has also been used to statically predict the
outcome of branches~\cite{Patterson95}, to detect buffer overflow
vulnerabilities~\cite{Simon08,Wagner00}, to find the trip count of
loops~\cite{Lokuciejewski09}
and even to synthesize hardware~\cite{Cong05,Lhairech10,Mahlke01}.

% The problem: few implementations, not extensive experimental reports.
Given this great importance, it comes as no surprise that the compiler
literature is rich in works describing in details algorithmic variations of
range analyses~\cite{Gawlitza09,Mahlke01,Stephenson00,Su05}.
On the other hand, none of these authors provide experimental evidence that
their approaches are able to deal with very large programs.
There are researchers who have implemented range analyses that scale up to
large programs~\cite{Patterson95,Blanchet03,Venet04}; nevertheless, because the
algorithm itself is not the main focus of their works, they neither give
details about their design choices nor provide experimental data about it.
This scenario was recently changed by Oh {\em et al.}~\cite{Oh12}, who
introduced an abstract interpretation framework which processes programs with
hundreds of thousands of lines of code.
Nevertheless, Oh {\em et al.} have designed a simple range analysis,
which does not handle comparisons between variables, for instance.
They do not discuss the precision of their implementation, but only its
runtime and memory consumption.
In this paper we claim to push this discussion considerably further.

% Our contribution.
In this paper we provide a complete description of a range analysis algorithm,
and show extensive experimental data that justifies our engineering choices.
Our first algorithmic contribution on top of previous works is a three-phases
approach to handle comparisons between variables without resorting to any
exponential time technique.
The few publicly available implementations of range analyses that we are
aware of, such as those in FLEX~\footnote{The MIT's FLEX/Harpoon compiler
provides an implementation of Stephenson's algorithm~\cite{Stephenson00}, and
is available at \texttt{http://flex.cscott.net/Harpoon/}.} and
gcc~\footnote{Gcc's VRP pass (at \texttt{http://gcc.gnu.org/svn/gcc/trunk/gcc/tree-vrp.c}) implements a variant of Patterson's
algorithm~\cite{Patterson95}.} only deal with comparisons between variables and
constants.
Even theoretical works, such as Su and Wagner's~\cite{Su05} or Gawlitza
{\em et al.}'s~\cite{Gawlitza09} suffer from this limitation.
This deficiency is one of the reasons explaining why none of these works has
made their way into industrial-strength compilers.
Two other insights allow our implementation to scale up to very large
programs.
We use Bodik's Extended Static Single Assignment (e-SSA) form~\cite{Bodik00} to
perform path-sensitive range analysis sparsely.
This program representation ensures that the interval associated with
a variable is constant along its entire live range.
Finally, we process the strongly connected components that underline our
constraint system in topological order.
It is well-known that this technique is essential to speedup constraint solving
algorithms~\cite[Sec 6.3]{Nielson99}; however, due to our three-phases
approach, a careful propagation of information along strong components not only
gives us speed, but also improves the precision of our results.

@@

We use the notation $\fun{ft}(v)$ to denote the {\em future} bounds of a
variable.
As we will show in Section~\ref{sub:micro}, once the growth pattern of $v$ is
known, we can replace $\fun{ft}(v)$ by an actual value.
After splitting the live ranges according to the rules stated above, we might
have to insert $\phi$-functions into the transformed program to re-convert it to
SSA form.
This last step avoids that two different names given to the same original
variable be simultaneously {\em alive} at the program code.
A variable $v$ is alive at a program point $p$ if the program's control flow
graph contains a path from $p$ to a site where $v$ is used, that does not
go across any re-definition of $v$.
Figure~\ref{fig:ex_standard_eSSA}(a) shows our running example changed into
standard e-SSA form.
We have not created variable names for $i_f$ and $j_f$, because neither
$i_1$ nor $i_j$ are dominated by the target of the conditional's else case.
In this example, new $\phi$-functions are not necessary: new variable
names are not alive together with the original variables.
The part (b) of this figure shows the solution that we get to this new
program.
The e-SSA form allows us to bind interval information directly to the live
ranges of variables; thus, giving us the opportunity to solve range analysis
sparsely.
More traditional approaches, which we call {\em dense analyses}, bind
interval information to pairs formed by variables and program points.

@@

\noindent
\textbf{Fixing futures.}
The ranges found by the growth analysis tells us which variables have fixed
bounds, independent on the intersections in the constraint system.
Thus, we can use actual limits to replace intersections bounded by futures.
Figure~\ref{fig:fix_intersects} shows the rules to perform these substitutions.
In order to correctly replace a future $\fun{ft}(v)$ that limits a variable
$v'$, we need to have already applied the growth analysis onto $v$.
Had we considered only data dependence edges, then it would be possible
that $N_v'$ be analyzed before $N_v$.
However, because of control dependence edges, this case cannot happen.
The control dependence edges ensure that any topological ordering of the
constraint graph either places $N_v$ before $N_{v'}$, or places these nodes
in the same strongly connected component.
For instance, in Figure~\ref{fig:ex_graph}, variables $j_1$ and $i_t$ are in
the same SCC only because of the control dependence edges.

@@

\noindent
\textbf{Precision.}
Our implementation of range analysis is remarkably precise, considering its
runtime.
Lakhdar {\em et al.}'s relational analysis~\cite{Lakhdar11}, for instance, takes
about 25 minutes to go over a program with almost 900 basic blocks.
We analyze programs of similar size in less than one second.
We do not claim our approach is as precise as such algorithms, even though we
are able to find exact bounds to 4/5 of the examples presented
in~\cite{Lakhdar11}.
On the contrary, this paper presents a compromise between precision and speed
that scales to very large programs.
Nevertheless, our results are far from being trivial.
We have implemented a dynamic profiler that measures, for each variable,
its upper and lower limits, given an execution of the target program.
Figure~\ref{fig:precision} compares our results with those measured
dynamically for the Stanford benchmark suite, which is publicly
available~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}}.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\textwidth]{images/precUpperBound}
\includegraphics[width=\textwidth]{images/precLowerBound}
\end{center}
\caption{\label{fig:precision}
(Upper) Comparison between static range analysis and dynamic profiler for
upper bounds.
(Lower) Comparison between static range analysis and dynamic profiler for
lower bounds. The numbers above the benchmark names give the number of
variables in each program.}
\end{figure}

@@

As we see in Figure~\ref{fig:precision}, 54.11\% of the lower limits that
we have estimated statically are exact.
Similarly, 51.99\% of our upper bounds are also tight.
The figure also shows that, on average, 37.39\% of our lower limits are
imprecise, and 35.40\% of our upper limits are imprecise.
This result is on pair with those obtained by more costly analysis, such as
Stephenson {\em et al.}'s~\cite{Stephenson00}.
However, whereas that approach have not been used with programs larger than
the Stanford benchmark suite, we, as shown before, have been able to
deal with remarkably larger programs.

@@

The strong components improve the precision of our growth analysis.
According to Figure~\ref{fig:impactSCC}, in some cases, as in \texttt{bzip2},
strong components increase our precision by 40\%.
The gains in precision happen because, by completely resolving a component,
we are able to propagate constant intervals to the next components, instead
of propagating intervals that can grow in both directions.
An an example, in Figure~\ref{fig:ex_partition_grow_crop} we pass the range
$[0, 99]$ from variable $k$ to the component that contains variable $j$.
Had we run the analysis in the entire constraint graph, by the time we
applied the growth analysis on $j$ we would still find $k$ bound to
$0, +\infty$.

@@

Figure~\ref{fig:impactESSA} compares these two program representations in
terms of runtime.
As we see in Figure~\ref{fig:impactESSA}(Left), the e-SSA form slows down our
analysis.
In some cases, as in \texttt{xalancbmk}, this slowdown increases execution
time by 71\%.
Runtime increases because the e-SSA form programs are larger than the SSA
form programs, as we show in Figure~\ref{fig:impactESSA}(Right).
The increase in size is due to the extra instructions used to partition the
live ranges of variables after conditionals.
However, this growth is small: in none of the integer programs in SPEC
CPU 2006 we verified an increase in code size of more than 9\%.
But, if the e-SSA form slowdowns the analysis runtime, its gains in precision
are remarkable, as seen in Figure~\ref{fig:precESSA}.
These gains happen because the e-SSA format lets the analysis to use the
results of conditional tests to narrow the ranges of variables.

@@

Figure~\ref{fig:wholeImpact} compares the precision of the intra and
inter-procedural analyses for the five largest programs in three different
categories of benchmarks: SPEC CPU 2006, the Stanford Suite~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}} and
Bitwise~\cite{Stephenson00}.
Our results for the SPEC programs were disappointingly: on the average for
the five largest programs, the intra-procedural version of our analysis saves
5.23\% of bits per variable.
The inter-procedural version increases this number to 8.89\%.
A manual inspection of the SPEC programs reveals that this result is expected:
these programs manipulate files, and their source codes do no provide enough
explicit constants to power our analysis up.
However, with numerical benchmarks we fare much better.
On the average our inter-procedural algorithm reduces the bitwidth of the
Stanford benchmarks by 36.24\%.
For Bitwise we obtain a bitwidth reduction of 12.27\%.
However, this average is lowered by two outliers: \texttt{edge\_detect} and
\texttt{sha}, which have been purposely engineered to be resilient against
range analyses~\cite{Stephenson00}.
The bitwise benchmarks were implemented by Stephenson
{\em et al.}~\cite{Stephenson00} to validate their intra-procedural bitwidth
analysis.
Our results are on par with those found by the original authors.
The bitwise programs contain only the \texttt{main} function; thus, different
versions of our algorithm find the same results when applied onto these
programs.

@@

Another way to increase the precision of range analysis is via a
context-sensitive implementation.
Context-sensitiveness allows us to distinguish different calling sites of
the same function.
Figure~\ref{fig:context} shows why the ability to make this distinction is
important for precision.
In Figure~\ref{fig:context}(a) we have two different calls of function
\texttt{foo}.
If we apply the trivial intra-procedural approach of Section~\ref{sub:whole},
then we get the graph shown in Figure~\ref{fig:context}(b).
In other words, if a function is called more than once, then its formal
parameters will receive information from many actual parameters.
We use $\phi$-functions to bind this information together into a single
flow.
However, in this case the multiple assignment of values to parameters
makes the ranges of these parameters very large, whereas in reality they
are not.
A way to circumvent this source of imprecision is via function inlining,
as we show in Figure~\ref{fig:context}(c).
The results that we can derive for the transformed program are more
precise, as each input parameter is assigned a single value.

@@

Figure~\ref{fig:wholeImpact} also shows how function inlining modifies the
precision of our results.
It is difficult to find an adequate way to compare the precision of
our analysis with, and without inlining.
This difficulty stems from the fact that this transformation tends to change
the target program too much.
In absolute numbers, we always reduce the bitwidth of more variables after
function inlining.
However, proportionally function inlining leads to a smaller percentage of
bitwidth reduction for many benchmarks.
In the Stanford Collection, for instance, where most of the functions are
called in only one location, inlining leads to worst precision results.
On the other hand, for the SPEC programs, inlining, even in terms of
percentage of reduction, tends to increase our measure of precision.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\textwidth]{images/wholeImpact}
\end{center}
\caption{\label{fig:wholeImpact}
The impact of whole program analysis on precision. Each bar shows precision in
\%bitwidth reduction.}
\end{figure}

\noindent
\textbf{Intra vs Inter-procedural runtimes.}
Figure~\ref{fig:timeComp}(Right) compares three different execution modes.
Bars are normalized to the time to run the intra-procedural analysis
without inlining.
On the average, the intra-procedural mode is 28.92\% faster than the
inter-procedural mode.
If we perform function inlining, then this difference is 45.87\%.
These numbers are close because our runtime is bound to the
size of the strong components.
Although function inlining can increase the number of strongly connected
components in the constraint graph, it cannot increase the size of the largest
component, when compared to the simple inter-procedural analysis of
Section~\ref{sub:whole}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By combining the different widening operators -- simple or jump-set -- with
the different number of pre-evaluations -- in our case 0 or 16, we have four
different widening strategies.
Figure~\ref{fig:timeComp}(Right) compares the runtime of all these strategies
for the integer programs in the SPEC CPU 2006 collection.
We have observed no measurable different between the jump-set and the simple
operator: the latter is 0.93\% faster, e.g., 1.0093x faster.
The strategy that precedes the simple operator with 16 rounds of pre-evaluation
is 7\% slower than the strategy that does not do any pre-evaluation.
Finally, the combination of 16 rounds of pre-evaluation, plus jump-set widening
is 13\% slower than the simple widening strategy.
We observe an anomalous behavior in \texttt{astar} and \texttt{mcf}: the simple
strategy results in a small slowdown.
These benchmarks have the two fastest runtimes in the benchmark suite; e.g.,
we analyze \texttt{mcf} in 0.02 seconds.
Thus, we believe that the unexpected behavior is due to runtime noise that
outlives a sequence of 100 executions of each benchmark.

@@

Most eminent among these representations is the Static Single Assignment
form~\cite{Cytron91}, which suits well forward flow analyses, such as reaching
definitions.
Since its debut, the SSA form has been expanded in different ways.
For instance, the Gated SSA form allows the static association of logical
predicates with data-flow paths~\cite{Ottenstein90,Tu95}.
Scott Ananian has introduced in the late nineties the {\em Static Single
Information} (SSI) form, a program representation that supports both forward and
backward analyses~\cite{Ananian99}.
This representation was later revisited by Jeremy Singer~\cite{Singer06} and,
a few years later, by Boissinot {\em et al.}~\cite{Benoit09}.
Singer provided new algorithms plus examples of applications that benefit from
the SSI form, and Boissinot {\em et al.}, in an effort to clarify some
misconceptions about this program representation, introduced the notions of
{\em week} and {\em strong} SSA form.
Another important representation, which supports data-flow analyses that
acquire information from uses, is the \emph{Static Single Use} form (SSU).
There exists many variants of
SSU~\cite{Plevyak96,George03,Lo98}.
For instance, the ``strict'' SSU form enforces that each definition reaches a
single use, whereas SSI and other variations of SSU allow two consecutive uses
of a variable on the same path.
The program representation that we have used in this paper -- the {\em Extended
Static Single Assignment} (e-SSA) form -- was introduced by Bodik
{\em et al.}~\cite{Bodik00}.

>>>>

% The context: why range analysis is necessary.
The analysis of integer variables on the interval lattice has been the
canonical example of abstract interpretation since its introduction in
Cousot and Cousot's seminal paper~\cite{Cousot77}.
Compilers use range analysis to infer the possible values that discrete
variables may assume during program execution.
This analysis has many uses.
For instance, it allows the optimizing compiler to remove from the program text
redundant overflow tests~\cite{Sol11} and unnecessary array bound
checks~\cite{Bodik00,Gampe11}.
Furthermore, range analysis is essential to bitwidth aware register
allocators~\cite{Barik06,Tallam03}, register allocators that handle registers of
different sizes~\cite{Kong98,Pereira08,Scholz02}, and scratchpad cache
allocators~\cite{Yanqin11}.
Additionally, range analysis has also been used to statically predict the
outcome of branches~\cite{Patterson95}, to detect buffer overflow
vulnerabilities~\cite{Simon08,Wagner00}, to find the trip count of
loops~\cite{Lokuciejewski09}
and even to synthesize hardware~\cite{Cong05,Lhairech10,Mahlke01}.

% The problem: few implementations, not extensive experimental reports.
Given this great importance, it comes as no surprise that the compiler
literature is rich in works describing in details algorithmic variations of
range analyses~\cite{Mahlke01,Gawlitza09,Stephenson00,Su05}.
On the other hand, none of these authors provide experimental evidence that
their approaches are able to deal with very large programs.
There are researchers who have implemented range analyses that scale up to
large programs~\cite{Patterson95,Blanchet03,Venet04}; nevertheless, because the
algorithm itself is not the main focus of their works, they neither give
details about their design choices nor provide experimental data about it.
This scenario was recently changed by Oh {\em et al.}~\cite{Oh12}, who
introduced an abstract interpretation framework which processes programs with
hundreds of thousands of lines of code.
Nevertheless, Oh {\em et al.} have designed a simple range analysis,
which does not handle comparisons between variables, for instance.
They do not discuss the precision of their implementation, but only its
runtime and memory consumption.
In this paper we claim to push this discussion considerably further.

% Our contribution.
In this paper we provide a complete description of a range analysis algorithm,
and show extensive experimental data that justifies our engineering choices.
Our first algorithmic contribution on top of previous works is a three-phases
approach to handle comparisons between variables without resorting to any
exponential time technique.
The few publicly available implementations of range analyses that we are
aware of, such as those in FLEX~\footnote{The MIT's FLEX/Harpoon compiler
provides an implementation of Stephenson's algorithm~\cite{Stephenson00}, and
is available at \texttt{http://flex.cscott.net/Harpoon/}.},
gcc~\footnote{Gcc's VRP pass (at \texttt{http://gcc.gnu.org/svn/gcc/trunk/gcc/tree-vrp.c}) implements a variant of Patterson's
algorithm~\cite{Patterson95}.} or Mozilla's IonMonkey~\footnote{\texttt{hg.mozilla.org/projects/ionmonkey/file/629d1e6251b9/js/src/ion/RangeAnalysis.cpp}} only deal with comparisons between variables and
constants.
Even theoretical works, such as Su and Wagner's~\cite{Su05} or Gawlitza
{\em et al.}'s~\cite{Gawlitza09} suffer from this limitation.
This deficiency is one of the reasons explaining why none of these works has
made their way into industrial-strength compilers.
Two other insights allow our implementation to scale up to very large
programs.
We use Bodik's Extended Static Single Assignment (e-SSA) form~\cite{Bodik00} to
perform path-sensitive range analysis sparsely.
This program representation ensures that the interval associated with
a variable is constant along its entire live range.
Finally, we process the strongly connected components that underline our
constraint system in topological order.
It is well-known that this technique is essential to speedup constraint solving
algorithms~\cite[Sec 6.3]{Nielson99}; however, due to our three-phases
approach, a careful propagation of information along strong components not only
gives us speed, but also improves the precision of our results.

@@

We use the notation $\fun{ft}(v)$ to denote the {\em future} bounds of a
variable.
As we will show in Section~\ref{sub:micro}, once the growth pattern of $v$ is
known, we can replace $\fun{ft}(v)$ by an actual value.
After splitting the live ranges according to the rules stated above, we might
have to insert $\phi$-functions into the transformed program to re-convert it to
SSA form.
This last step avoids that two different names given to the same original
variable be simultaneously {\em alive} at the program code.
A variable $v$ is alive at a program point $p$ if the program's control flow
graph contains a path from $p$ to a site where $v$ is used, that does not
go across any re-definition of $v$.
Figure~\ref{fig:ex_standard_eSSA}(a) shows our running example changed into
standard e-SSA form.
We have not created variable names for $i_f$ and $j_f$, because neither
$i_1$ nor $j_1$, the variables that have been renamed, are dominated by the
target of the conditional's else case.
In this example, new $\phi$-functions are not necessary: new variable
names are not alive together with the original variables.
The part (b) of this figure shows the solution that we get to this new
program.
The e-SSA form allows us to bind interval information directly to the live
ranges of variables; thus, giving us the opportunity to solve range analysis
sparsely.
More traditional approaches, which we call {\em dense analyses}, bind
interval information to pairs formed by variables and program points.

@@

\noindent
\textbf{Fixing futures.}
The ranges found by the growth analysis tells us which variables have fixed
bounds, independent on the intersections in the constraint system.
Thus, we can use actual limits to replace intersections bounded by futures.
Figure~\ref{fig:fix_intersects} shows the rules to perform these substitutions.
In order to correctly replace a future $\fun{ft}(v)$ that limits a variable
$v'$, we need to have already applied the growth analysis onto $v$.
Had we considered only data dependence edges, then it would be possible
that $N_{v'}$ be analyzed before $N_v$.
However, because of control dependence edges, this case cannot happen.
The control dependence edges ensure that any topological ordering of the
constraint graph either places $N_v$ before $N_{v'}$, or places these nodes
in the same strongly connected component.
For instance, in Figure~\ref{fig:ex_graph}, variables $j_1$ and $i_t$ are in
the same SCC only because of the control dependence edges.

@@

\noindent
\textbf{Precision.}
Our implementation of range analysis is remarkably precise, considering its
runtime.
Lakhdar {\em et al.}'s relational analysis~\cite{Lakhdar11}, for instance, takes
about 25 minutes to go over a program with almost 900 basic blocks.
We analyze programs of similar size in less than one second.
We do not claim our approach is as precise as such algorithms, even though we
are able to find exact bounds to 4/5 of the examples presented
in~\cite{Lakhdar11}.
On the contrary, this paper presents a compromise between precision and speed
that scales to very large programs.
Nevertheless, our results are far from being trivial.
We have implemented a dynamic profiler that measures, for each variable,
its upper and lower limits, given an execution of the target program.
Figure~\ref{fig:precision} compares our results with those measured
dynamically for the Stanford benchmark suite, which is publicly
available~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}}.
We chose Stanford because these benchmarks do not read data from external
files; hence, imprecisions are due to library functions that we cannot
analyze.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.97\textwidth]{images/precUpperBound}
\includegraphics[width=\textwidth]{images/precLowerBound}
\end{center}
\caption{\label{fig:precision}
(Upper) Comparison between static range analysis and dynamic profiler for
upper bounds.
(Lower) Comparison between static range analysis and dynamic profiler for
lower bounds. The numbers above the benchmark names give the number of
variables in each program.}
\end{figure}

@@

As we see in Figure~\ref{fig:precision}, 54.11\% of the lower limits that
we have estimated statically are exact.
Similarly, 51.99\% of our upper bounds are also tight.
The figure also shows that, on average, 37.39\% of our lower limits are
imprecise, and 35.40\% of our upper limits are imprecise.
This result is on pair with those obtained by more costly analysis, such as
Stephenson {\em et al.}'s~\cite{Stephenson00}.
However, whereas that approach has only been tested on single functions,
we have been able to deal with remarkably larger programs.

@@

The strong components improve the precision of our growth analysis.
According to Figure~\ref{fig:impactSCC}, in some cases, as in \texttt{bzip2},
strong components increase our precision by 40\%.
The gains in precision happen because, by completely resolving a component,
we are able to propagate constant intervals to the next components, instead
of propagating intervals that can grow in both directions.
As an example, in Figure~\ref{fig:ex_partition_grow_crop} we pass the range
$[0, 99]$ from variable $k$ to the component that contains variable $j$.
Had we run the analysis in the entire constraint graph, by the time we
applied the growth analysis on $j$ we would still find $k$ bound to
$0, +\infty$.

@@

Figure~\ref{fig:impactESSA} compares these two program representations in
terms of runtime.
As we see in Figure~\ref{fig:impactESSA}(Left), the e-SSA form slows down our
analysis.
In some cases, as in \texttt{xalancbmk}, this slowdown increases execution
time by 71\%.
Runtime increases for two reasons.
Firstly, the e-SSA form programs are larger than the SSA form programs, as we
show in Figure~\ref{fig:impactESSA}(Right).
However, this growth is small: in none of the integer programs in SPEC
CPU 2006 we verified an increase in code size of more than 9\%.
Secondly, the e-SSA form program has futures; hence requiring the future
resolution phase of our algorithm, which is not necessary in SSA form programs.
Nevertheless, if the e-SSA form slowdowns the analysis runtime, its gains in precision
are remarkable, as seen in Figure~\ref{fig:precESSA}.
These gains happen because the e-SSA format lets the analysis to use the
results of conditional tests to narrow the ranges of variables.

@@

Figure~\ref{fig:wholeImpact} compares the precision of the intra and
inter-procedural analyses for the five largest programs in three different
categories of benchmarks: SPEC CPU 2006, the Stanford Suite~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}} and
Bitwise~\cite{Stephenson00}.
Our results for the SPEC programs were disappointing: on the average for
the five largest programs, the intra-procedural version of our analysis saves
5.23\% of bits per variable.
The inter-procedural version increases this number to 8.89\%.
A manual inspection of the SPEC programs reveals that this result is expected:
these programs manipulate files, and their source codes do not provide enough
explicit constants to power our analysis up.
However, with numerical benchmarks we fare much better.
On the average our inter-procedural algorithm reduces the bitwidth of the
Stanford benchmarks by 36.24\%.
For Bitwise we obtain a bitwidth reduction of 12.27\%.
However, this average is lowered by two outliers: \texttt{edge\_detect} and
\texttt{sha}, which have been purposely engineered to be resilient against
range analyses~\cite{Stephenson00}.
The bitwise benchmarks were implemented by Stephenson
{\em et al.}~\cite{Stephenson00} to validate their intra-procedural bitwidth
analysis.
Our results are on par with those found by the original authors.
The bitwise programs contain only the \texttt{main} function; thus, different
versions of our algorithm find the same results when applied onto these
programs.

@@

Another way to increase the precision of range analysis is via a
context-sensitive implementation.
Context-sensitiveness allows us to distinguish different calling sites of
the same function.
Figure~\ref{fig:context} shows why the ability to make this distinction is
important for precision.
In Figure~\ref{fig:context}(a) we have two different calls of function
\texttt{foo}.
If we apply the trivial inter-procedural approach of Section~\ref{sub:whole},
then we get the graph shown in Figure~\ref{fig:context}(b).
In other words, if a function is called more than once, then its formal
parameters will receive information from many actual parameters.
We use $\phi$-functions to bind this information together into a single
flow.
However, in this case the multiple assignment of values to parameters
makes the ranges of these parameters very large, whereas in reality they
are not.
A way to circumvent this source of imprecision is via function inlining,
as we show in Figure~\ref{fig:context}(c).
The results that we can derive for the transformed program are more
precise, as each input parameter is assigned a single value.

@@

Figure~\ref{fig:wholeImpact} also shows how function inlining modifies the
precision of our results.
It is difficult to find an adequate way to compare the precision of
our analysis with, and without inlining.
This difficulty stems from the fact that this transformation tends to change
the target program too much.
In absolute numbers, we always reduce the bitwidth of more variables after
function inlining.
However, proportionally function inlining leads to a smaller percentage of
bitwidth reduction for many benchmarks.
In the Stanford Collection, for instance, where most of the functions are
called in only one location, inlining leads to worse precision results.
On the other hand, for the SPEC programs, inlining, even in terms of
percentage of reduction, tends to increase our measure of precision.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\textwidth]{images/wholeImpact}
\end{center}
\caption{\label{fig:wholeImpact}
The impact of inter-procedural analysis on precision. Each bar shows precision
in \%bitwidth reduction.}
\end{figure}

\noindent
\textbf{Intra vs Inter-procedural runtimes.}
Figure~\ref{fig:timeComp}(Left) compares three different execution modes.
Bars are normalized to the time to run the intra-procedural analysis
without inlining.
On the average, the intra-procedural mode is 28.92\% faster than the
inter-procedural mode.
If we perform function inlining, then this difference is 45.87\%.
These numbers are close because our runtime is bound to the
size of the strong components.
Although function inlining can increase the number of strongly connected
components in the constraint graph, it cannot increase the size of the largest
component, when compared to the simple inter-procedural analysis of
Section~\ref{sub:whole}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By combining the different widening operators -- simple or jump-set -- with
the different number of pre-evaluations -- in our case 0 or 16, we have four
different widening strategies.
Figure~\ref{fig:timeComp}(Right) compares the runtime of all these strategies
for the integer programs in the SPEC CPU 2006 collection.
We have observed no measurable difference between the jump-set and the simple
operator: the latter is 0.93\% faster, e.g., 1.0093x faster.
The strategy that precedes the simple operator with 16 rounds of pre-evaluation
is 7\% slower than the strategy that does not do any pre-evaluation.
Finally, the combination of 16 rounds of pre-evaluation, plus jump-set widening
is 13\% slower than the simple widening strategy.
We observe an anomalous behavior in \texttt{astar} and \texttt{mcf}: the simple
strategy results in a small slowdown.
These benchmarks have the two fastest runtimes in the benchmark suite; e.g.,
we analyze \texttt{mcf} in 0.02 seconds.
Thus, we believe that the unexpected behavior is due to runtime noise that
outlives a sequence of 100 executions of each benchmark.

@@

Most eminent among these representations is the Static Single Assignment
form~\cite{Cytron91}, which suits well forward flow analyses, such as reaching
definitions.
Since its debut, the SSA form has been expanded in different ways.
For instance, the Gated SSA form allows the static association of logical
predicates with data-flow paths~\cite{Ottenstein90,Tu95}.
Scott Ananian has introduced in the late nineties the {\em Static Single
Information} (SSI) form, a program representation that supports both forward and
backward analyses~\cite{Ananian99}.
This representation was later revisited by Jeremy Singer~\cite{Singer06} and,
a few years later, by Boissinot {\em et al.}~\cite{Benoit09}.
Singer provided new algorithms plus examples of applications that benefit from
the SSI form, and Boissinot {\em et al.}, in an effort to clarify some
misconceptions about this program representation, introduced the notions of
{\em weak} and {\em strong} SSI form.
Another important representation, which supports data-flow analyses that
acquire information from uses, is the \emph{Static Single Use} form (SSU).
There exists many variants of
SSU~\cite{Plevyak96,George03,Lo98}.
For instance, the ``strict'' SSU form enforces that each definition reaches a
single use, whereas SSI and other variations of SSU allow two consecutive uses
of a variable on the same path.
The program representation that we have used in this paper -- the {\em Extended
Static Single Assignment} (e-SSA) form -- was introduced by Bodik
{\em et al.}~\cite{Bodik00}.
